Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-10-24 response_to_reviewers.Rmd

> Table 2 seems unnecessary, as the content is entirely summarized in the text.

We have removed Table 2.

> After the comprehensive views in Sections 2 and 3, Section 4 seems sparse on details. Much detail is given on the Bradley-Terry model, but only one example is given of a possible usage. Has this been used in other sports? Has it changed how we think about those sports or team strategies? An explanation of Elo rating should be given; I for one have no knowledge of how to compute it.

We thank the reviewer for this suggestion. We have added a subsection on Elo rating, with a mathematical defination followed by a calculation example.

@@

> 12, 23: This is not much of a case study.  Perhaps some additional information on the data available, ways it could be used to calculate game state expected valued, win probability, player ratings….

XX Yeah

> 12, 37: Theming *teamcolors* is a great *R* package, not sure it is part of the "big ideas" in sports analytics.  A single line in the first paragraph of this section would seem to be better.

We have added examples from the `SportyR` package, which draws playing surfaces, to help flesh out this subsection. We also replaced the previous Figure 6 with a new Figure 5 that illustrates how the use of team colors is particularly impactful for sports-releated data graphics.

### 2022-10-24 sports_analytics.Rmd
close #38

Each professional sports team has its own brand, most obviously identified by a team logo and set of colors. 
The \pkg{teamcolors} R package [@R-teamcolors] provides color palettes and logos for men's and women's professional and collegiate sports teams, as well as color and fill scale functions compatible with \pkg{ggplot2} [@R-ggplot2].
For example, the NFL teams' colors and logos shown in Figure \ref{fig:nflfastr-wp} were provided by the \pkg{teamcolors} package. 
Figure \ref{fig:teamcolors} illustrates the full array of team color palettes for professional sports leagues.

<!-- 
```{r teamcolors, warning=FALSE, fig.cap="Professional sports team color palettes provided by the \\pkg{teamcolors} package. ", fig.height=9}
teamcolors::show_team_col() 
``` 
-->

Sharing a similar goal with \pkg{teamcolors} is \pkg{nflplotR} [@R-nflplotR], which also provides \pkg{ggplot2} extensions but is designed specifically for the NFL.
A great feature of \pkg{nflplotR} is the \pkg{ggplot2} "geom" (geometric object) functions that enhance high-quality plotting of NFL team logos and player images.
Figure \ref{fig:nflplotr} shows a scatterplot of offensive and defensive expected points added for NFL teams in the 2021 regular season.
The logos of all 32 American football clubs are plotted in place of the usual dots, giving a better description of which team each data point represents.

>>>>

Each professional sports team has its own brand, most obviously identified by a team logo and set of colors. 
The \pkg{teamcolors} R package [@R-teamcolors] provides color palettes and logos for men's and women's professional and collegiate sports teams, as well as color and fill scale functions compatible with \pkg{ggplot2} [@R-ggplot2].
For example, the NFL teams' colors and logos shown in Figure \ref{fig:nflfastr-wp} were provided by the \pkg{teamcolors} package. 
Figure \ref{fig:mlb_standings} illustrates how the use of team colors, which have a natural association for many sports fans, can help to untangle what would otherwise be messy data graphics. 
In Figure \ref{fig:mlb_standings}, 30 different lines are plotted on top of one another, crisscrossing and intersecting in various unpredictable ways. 
However, the use of team colors to identify the lines makes it possible to follow the trajectory of most teams over the course of the season.

```{r}
# remotes::install_github("beanumber/retro")
library(retrosheet)
retro_gl <- get_retrosheet("game", 2021)
games_2021 <- retro_gl |>
  mutate(home_win = HmRuns > VisRuns) %>%
  select(Date, VisTm, HmTm, home_win) %>%
  pivot_longer(
    -c(Date, home_win),
    names_to = "is_home", values_to = "team_id", 
  ) %>%
  mutate(
    is_win = ifelse(
      is_home == "HmTm", home_win, abs(1 - home_win))
  ) %>%
  group_by(team_id) %>%
  mutate(
    cum_W = cumsum(is_win), 
    cum_L = cumsum(!is_win), 
    cum_G = cum_W + cum_L, 
    G_over_500 = cum_W - cum_L, 
    last_game = cum_G == max(cum_G)
  )
retro_teams <- retrosheet::getTeamIDs(2021) |>
  enframe(name = "mascot", value = "retro_id")
mlb_pal <- teamcolors |>
  filter(league == "mlb", str_detect(division, "NL")) |>
  inner_join(retro_teams, by = "mascot") |>
  select(retro_id, primary) |>
  deframe()
games_2021 <- games_2021 |>
  filter(team_id %in% names(mlb_pal))
```

```{r mlb_standings, fig.cap="The progression of National Leauge team standings during the 2021 Major League Baseball season. Note how the use of team colors makes it possible to untangle what would otherwise be a messy jumble of indistinguishable lines. Data provided by \\pkg{retrosheet} and colors provided by \\pkg{teamcolors}."}
ggplot(
    games_2021, 
    aes(x = cum_G, y = G_over_500, color = team_id)
  ) + 
  geom_hline(yintercept = 18, color = "darkgray", lty = 2) + 
  geom_line() +  
  ggrepel::geom_label_repel(
    data = filter(games_2021, last_game), 
    aes(label = team_id)
  ) + 
  scale_color_manual(
    guide = "none", name = "Team",
    values = mlb_pal
  ) +
  scale_y_continuous("Games above .500") + 
  scale_x_continuous("Cumulative Games Played") + 
  labs(title = "Daily MLB Standings, 2021", 
       caption = "Source: Retrosheet")
```

\pkg{nflplotR} [@R-nflplotR] has a similar goal to \pkg{teamcolors}.
It also provides \pkg{ggplot2} extensions but is designed specifically for the NFL.
A great feature of \pkg{nflplotR} is the collection of `geom_*()` (geometric object) functions that enhance high-quality plotting of NFL team logos and player images with \pkg{ggplot2}.
Figure \ref{fig:nflplotr} shows a scatterplot of offensive and defensive expected points added for NFL teams in the 2021 regular season.
The logos of all 32 American football clubs are plotted in place of the usual dots, making it easier for the reader to identify which team each data point represents.
